---
title: "Weather Effects"
author: "nthehai01"
date: "2/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.


## Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

- [Storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

- National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.


## Data Processing

#### Downloading data

Set up directory for downloading data

```{r}
dir.create("./data", showWarnings = FALSE)

downloadURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
downloadedFile <- "./data/repdata_data_StormData.csv.bz2"
```

Download and unzip data
```{r}
library(R.utils)
if(!file.exists(downloadedFile)) {
    download.file(downloadURL, downloadedFile, method = "curl")
    bunzip2(downloadedFile, destname = "./data/stormdata.csv", remove = FALSE)
}
```

#### Loading data

The actual dimension of the data is 902297x37. To reduce the time-consuming when loading it to the environment, we should read only the useful columns. Here are the useful columns and their column indexes:

| Column name | Column index |
|:-----------:|:------------:|
| EVTYPE      | 8            |
| FATALITIES  | 23           |
| INJURIES    | 24           |
| PROPDMG     | 25           |
| PROPDMGEXP  | 26           |
| CROPDMG     | 27           |
| CROPDMGEXP  | 28           |

To load these specific features, we need to specify their classes as ```numeric``` or ```character```, the types of other columns will be set to ```NULL```:
```{r, cache=TRUE}
stormdata <- read.csv("./data/stormdata.csv",
                      colClasses = c(rep("NULL", 7),
                                     "character",
                                     rep("NULL", 14),
                                     rep("numeric", 3),
                                     "character",
                                     "numeric",
                                     "character",
                                     rep("NULL", 9)),
                      sep = ",",
                      header = TRUE)
```

#### Data summary:

*Take a quick look of our data:*
```{r}
head(stormdata)
```
*Summary:*
```{r}
summary(stormdata)
```


#### Handle Exponent value of PROPDMGEXP and CROPDMGEXP

**These are possible values of PROPDMGEXP and CROPDMGEXP:**

- B or b = Billion = 10^9^ 
- M or m = Million = 10^6^
- K or k = Thousand = 10^3^
- H or h = Hundred = 10^2^
- The symbol "-" refers to less than = 10^0^ (ignore it)
- The symbol "+" refers to greater than = 10^0^ (ignore it)
- The symbol "?" refers to low certainty than = 10^0^ (ignore it)
- The number from 0 to 10 represent the power of ten = 10^TheNumber^
- The black/empty character = 10^0^

For more information, consider to visit [here](https://www.coursera.org/learn/reproducible-research/discussions/forums/FZGy9Sj0Eea8jw6UvTi2Tw/threads/38y35MMiEeiERhLphT2-QA) or [here](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html)

**1. Convert the Exponent columns to number:**

*Create a converter data frame for each symbol and its value:*
```{r}
symbol <- c("B", "b", "M", "m", "K", "k", "H", "h",
            "-", "+", "?", as.character(0:10), "")
value <- c(rep(10^9, 2), rep(10^6, 2), rep(10^3, 2), rep(10^2, 2), 
           rep(10^0, 3), 10^c(0:10), 10^0)

converter <- data.frame(Symbol = symbol, Value = value)
```

```{r}
head(converter)
```

*Replace each symbol by its value in the ```stormdata```:*
```{r, cache=TRUE}
## process for PROPDMGEXP
temp <- sapply(stormdata$PROPDMGEXP, 
                               function(ele) converter[converter$Symbol == ele, 2])
stormdata$PROPDMGEXP <- unlist(temp, use.names = FALSE)

## process for CROPDMGEXP
temp <- sapply(stormdata$CROPDMGEXP, 
                               function(ele) converter[converter$Symbol == ele, 2])
stormdata$CROPDMGEXP <- unlist(temp, use.names = FALSE)
```

```{r}
head(stormdata)
```

**2. Re-calculate the Property and Crop Damage in PROPDMG and CROPDMG columns:**

*Re-calculate:*
```{r, cache=TRUE}
stormdata$PROPDMG <- stormdata$PROPDMG * stormdata$PROPDMGEXP
stormdata$CROPDMG <- stormdata$CROPDMG * stormdata$CROPDMGEXP
```

```{r}
head(stormdata)
```

*Remove ```PROPDMGEXP``` and ```CROPDMGEXP``` columns:*
```{r}
stormdata$PROPDMGEXP <- NULL
stormdata$CROPDMGEXP <- NULL
```

```{r}
head(stormdata)
```


## Analysis

In this part, we will group the records that have same ```EVTYPE``` and then calculate the sum of each feature:
```{r}
library(dplyr)
group <- stormdata %>% 
    group_by(EVTYPE) %>%
    summarise_all(sum)
```

```{r}
head(group)
```


## Results

**Function to generate a bar plot:**

This function called ```generate.bar``` with input:

- ```df```: data frame
- ```x```: value for x-axis
- ```y```: value for y-axis
- ```x.lab```: label for x-axis
- ```y.lab```: label for y-axis

```{r}
library(ggplot2)
generate.bar <- function(df, x, y, x.lab, y.lab, title) {
    p <- ggplot(df, aes(x = x, y = y, fill = x))
    p <- p + geom_bar(stat = "identity") +
      xlab(x.lab) +
      ylab(y.lab) +
      ggtitle(title) +
      theme(legend.position = "none") +  ## remove legends
      theme(text = element_text(size = 12)) +  ## resize text
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))  ## rotate labels
}
```


#### Question 1: Across the United States, which types of events (as indicated in the ```EVTYPE``` variable) are most harmful with respect to population health?

**In this part, we will generate the bar plots to find the top 5 weather events that are most harmful to the US citizen for each type of damage:**

*In term of Fatalities damage:*

Extract the top 5 worst harmful events that result in Fatalities damage:
```{r}
fatalities <- group[order(-group$FATALITIES), ]
fatalities <- fatalities[1:5, ]
```

```{r}
head(fatalities)
```

Create a plot for this damage:
```{r}
fatalities.plot <- generate.bar(fatalities, 
                                fatalities$EVTYPE, fatalities$FATALITIES,
                                "Type of event", "Fatalities damage", 
                                "Top 5 most harmful events damaged in Fatalities")
```

*In term of Injuries damage:*

Extract the top 5 worst harmful events that result in Injuries damage:
```{r}
injuries <- group[order(-group$INJURIES), ]
injuries <- injuries[1:5, ]
```

```{r}
head(injuries)
```

Create a plot for this damage:
```{r}
injuries.plot <- generate.bar(injuries, 
                              injuries$EVTYPE, injuries$INJURIES,
                              "Type of event", "Injuries damage",
                              "Top 5 most harmful events damaged in Injuries")
```

*Plot both these types of damage into a panel*
```{r, fig.height=5, fig.width=10}
library(gridExtra)
grid.arrange(fatalities.plot, injuries.plot, ncol = 2)
```

#### Question 2: Across the United States, which types of events have the greatest economic consequences?

**In this part, we will generate the bar plots to find the top 5 weather events that are most harmful to the US citizen for each type of economic consequences:**

*In term of Property consequences:*

Extract the top 5 worst harmful events that result in Property consequences:
```{r}
property <- group[order(-group$PROPDMG), ]
property <- property[1:5, ]
```

```{r}
head(property)
```

Create a plot for this economic consequences:
```{r}
property.plot <- generate.bar(property, 
                              property$EVTYPE, property$PROPDMG,
                              "Type of event", "Property consequences",
                              "Top 5 most harmful events damaged in Property")
```

*In term of Crop consequences:*

Extract the top 5 worst harmful events that result in Crop consequences:
```{r}
crop <- group[order(-group$CROPDMG), ]
crop <- crop[1:5, ]
```

```{r}
head(crop)
```

Create a plot for this economic consequences:
```{r}
crop.plot <- generate.bar(crop, 
                          crop$EVTYPE, crop$CROPDMG,
                          "Type of event", "Injuries damage",
                          "Top 5 most harmful events damaged in Crop")
```

*Plot both these types of economic consequence into a panel*
```{r, fig.height=7, fig.width=10}
grid.arrange(property.plot, crop.plot, ncol = 2)
```